On the treatment of discordant data in detrital zircon U--Pb geochronology

### 2020-11-22 template.tex
First commit.

\usepackage{upquote}

@@

\title{On the treatment of discordant detrital zircon U--Pb data}

@@

\begin{abstract}
  Detrital zircon U--Pb geochronology is a staple of sedimentary
  provenance analysis and crustal evolution studies. Constructing
  detrital age spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This situation is expected to
  worsen in the future as a result of increasing analytical precision,
  which reveals ever smaller deviations from isotopic concordance.
  The analysis of discordant U--Pb data involves two largely arbitrary
  decisions.

@@

  Applying these six discordance filters to a 10,000-grain dataset of
  detrital zircon U--Pb compositions reveals that: (i) the relative
  age discordance filter tends to suppress the young ($<1.5$~Ga) age
  components in U--Pb age spectra, whilst inflating the older
  ($>1.5$~Ga) age components; (ii) the absolute age difference and
  Stacey-Kramers discordance filters are more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it yields nearly identical
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb age spectra.  All
  the methods presented in this paper have been implemented in the
  \texttt{IsoplotR} toolbox for geochronology.
\end{abstract}

@@

This paper addresses both of these issues. Section~\ref{sec:whichage}
shows that single-grain concordia ages \citep{ludwig1998} can be used
to avoid the arbitrary cutoff between the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb
methods. Section~\ref{sec:discordance1} compares and contrasts
existing discordance filters based on age disparity and p-values. It
shows that the p-value definition hurts both the accuracy and
precision of detrital geochronology. Section~\ref{sec:discordance2}
proposes three new ways to quantify discordance, based directly on
U--Pb compositions rather than the ages calculated therefrom. Two of
these new definitions are based on a logratio distance that is widely
used in compositional data analysis \citep{aitchison1986}. Finally,
Section~\ref{sec:application} applies the six discordance filters to a
compilation of detrital zircon U--Pb data. Although the true age
distribution of this dataset is unknowable, the results suggest that
the logratio based discordance filters produce the most accurate
results.

@@

The U--Pb method is based on three separate chronometers:
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb. The half-life of
\textsuperscript{235}U is more than six times shorter than that of
\textsuperscript{238}U, and \textsuperscript{235}U is more than 100
times less abundant than \textsuperscript{238}U. For these two
reasons, little \textsuperscript{207}Pb has been produced during the
last billion years of Earth history compared to
\textsuperscript{206}Pb. Consequently, the
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb methods are less
precise than the \textsuperscript{206}Pb/\textsuperscript{238}U method
during the Phanerozoic and Neoproterozoic.

However, during earlier stages of Earth's history,
\textsuperscript{235}U was significantly more abundant than it is
today. The \textsuperscript{238}U/\textsuperscript{235}U ratio was
$\sim$60 at 1Ga, $\sim$26 at 2Ga, $\sim$11 at 3Ga, and $\sim$5 at 4Ga.
Due to the greater abundance of \textsuperscript{235}U in this past,
and because it decays much faster than \textsuperscript{238}U, the
precision of the \textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb clocks exceeds that of
the \textsuperscript{206}Pb/\textsuperscript{238}U method during the
Palaeoproterozoic and Archaean. The gradual shift in sensitivity
between the two chronometers is visible in the slope of a
Tera-Wasserburg concordia line, which is steep at old ages (high
\textsuperscript{207}Pb/\textsuperscript{206}Pb gradient w.r.t. time)
and shallow at young ages (low
\textsuperscript{238}U/\textsuperscript{206}Pb gradient
w.r.t. time).

Most published detrital zircon U--Pb studies switch from
\textsuperscript{206}Pb/\textsuperscript{238}U to
\textsuperscript{207}Pb/\textsuperscript{206}Pb at some point during
the Proterozoic. Unfortunately there are two problems with such a
switch. First, it requires the selection of a discrete discordance
cutoff between the two methods. If this cutoff differs between two
studies (which it often does), then this complicates the
intercomparison of their respective age spectra. Second, the sudden
switch between the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb clocks is often marked
by a discrete step in the age spectrum \citep{puetz2018}. This step is
entirely artificial and obscures any geologically significant events
that might occur around the same time.

Both of these problems can be solved by using `hybrid' concordia ages
instead of `pure' \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages. Concordia ages
are defined by \citet{ludwig1998} as the `most likely' (in a
statistical sense) U--Pb age given the isotopic ratio composition and
its analytical uncertainty (Figure~\ref{fig:concordia}).  Let $r_{75}$
and $r_{68}$ be the measured
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{206}Pb/\textsuperscript{238}U ratios, respectively,
and let $\sigma[r_{68}]^2$, $\sigma[r_{76}]^2$,
$\sigma[r_{68},r_{76}]$ be their (co)variances.  Then the concordia
age $t_c$ is obtained by numerically minimising the sum of squares
$S$:
\begin{equation}
  S = \left[
    \begin{array}{@{}c@{}}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1
    \end{array}
    \right]^T\left[
    \begin{array}{@{}cc@{}}
      \sigma[r_{75}]^2 & \sigma[r_{75},r_{68}] \\
      \sigma[r_{75},r_{68}] & \sigma[r_{68}]^2
    \end{array}
    \right]^{-1}
  \left[
    \begin{array}{@{}c@{}}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1      
    \end{array}
    \right]
  \label{eq:S}
\end{equation}

@@

\section{Discordance filters: old definitions}\label{sec:discordance1}

@@

Detrital zircon U--Pb data can be filtered by removing all
measurements whose discordance values exceed a certain threshold
value. Typical cutoff values for $d_r$ are 10--30\%
\citep{gehrels2011}, whereas $d_p$ is generally set to 5\%
\citep{spencer2016}. Different discordance criteria produce different
U--Pb age spectra. For example, a relative age cutoff will
preferentially remove young grains whereas an absolute age cutoff is
more likely to remove old grains (Figure~\ref{fig:agediscordance}).

@@

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is 1\% discordant. If this grain were analysed by
LA-ICP-MS with an analytical precision of 2\%, say, then it would pass
the chi-square test and be accepted as being concordant. However, if
that same grain were analysed by TIMS with a precision of 0.2\%, then
the p-value criterion would reject it as being discordant. It seems
fundamentally unfair that an imprecise analytical method would be
favoured over a precise one (Figure~\ref{fig:TIMSvsLAICPMS}). This is
a pertinent problem because technical innovations are increasing the
precision of all analytical approaches to U--Pb geochronology.  As
precision improves, so does the ability to detect even small degrees
of discordance. Using the p-value criterion, there may come a time
when no detrital zircon passes this filter.

@@

\section{Discordance filters: new definitions}\label{sec:discordance2}

@@

\noindent where $[{}^{238}U/{}^{206}Pb]^\ast$ is the
\textsuperscript{238}U/\textsuperscript{206}Pb-ratio of the
intersection between concordia and a straight line connecting the
\textsuperscript{238}U/\textsuperscript{206}Pb--\textsuperscript{207}Pb/\textsuperscript{206}Pb
measurement to the inferred mantle composition (Figure~\ref{fig:SK}).

The common Pb definition of discordance is more forgiving for young
grains than it is for old ones. Importantly, if the discordance is
caused by common Pb, then the
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{206}Pb and concordia age
estimates are all negatively biased with respect to the true
age. However this bias can be removed by applying a common-Pb
correction \emph{after} the data have been filtered.

Each discordia definition that we have studied thus far is expressed
in different units. For the absolute age definition, degrees of
discordance are expressed in units of time (ranging from 0 to
4.5~Ga). The relative age definition uses fractions of time (ranging
from $-\infty$ to 1). The p-value definition expresses discordance in
terms of probability (ranging from 0 to 1). And the \citet{stacey1975}
definition uses fractions of ratios (ranging from $-\infty$ to
1). None of these scales is particularly intuitive or natural. They
certainly do not match the usual definition of \emph{distance} in the
geographical sense of the word.

To address this issue, it is useful to subject the U--Pb isotopic
ratio data to a logarithmic transformation. So instead of analysing
our data on a conventional Tera-Wasserburg concordia diagram, all
calculations can be done in
log(\textsuperscript{207}Pb/\textsuperscript{206}Pb) vs.
log(\textsuperscript{238}U/\textsuperscript{206}Pb) space. The
advantage of this transformation is that it produces values that are
free to range from $-\infty$ to $+\infty$. Within this infinite
dataspace, the Euclidean distance metric can be safely applied.

\begin{figure}
  \includegraphics[width=8.3cm]{Aitchison.pdf}
  \caption{Illustration of the two logratio distance definitions of
    discordance. $d_a$ is the perpendicular Aitchison distance from
    the measured logratio to the concordia line. $d_c$ is the
    Aitchison distance measured along a line connecting the measured
    value and the concordia composition.
  }
  \label{fig:aitchison}
\end{figure}

@@

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Era, when the U--Pb method is most reliable
(Figure~\ref{fig:dperp}).

The perpendicular Aitchison distance criterion does not take into
account the analytical precision of the isotopic measurements.  To
address this issue, we can also measure the Aitchison distance along a
line connecting the measured logratio and the maximum likelihood
composition on the concordia line:
\begin{equation}
  d_c = \mbox{sgn}[t_{76}-t_{68}] \sqrt{ dx(t_c)^2 + dy(t_c)^2 }
  \label{eq:da}
\end{equation}

@@

\begin{figure}
  \includegraphics[width=8.3cm]{TW-option-4.pdf}
  \caption{The $d_a$ criterion produces a parallel band to the
    (logarithmic) Tera-Wasserburg concordia line. The dashed line is
    plotted at an Aitchison distance of 0.1 from concordia, marking
    a 10\% discordance cutoff. The bottom panel presents the age
    discordance along this line. This shows that the $d_a$ criterion
    is lenient for Phanerozoic and Archaean age estimates, and more
    strict for Proterozoic grains.
  }
  \label{fig:dperp}
\end{figure}

@@

It is difficult to ascertain the mechanism causing discordance in any
particular zircon grain. Therefore, it is unclear which of the
definitions in Sections~\ref{sec:discordance1} and
\ref{sec:discordance2} is `correct'. All we can is do is apply the
methods to real samples and investigate their outcomes. This Section
will apply the six discordance filters to a compilation of 10,000
detrital zircon U--Pb analyses that were acquired at the London
Geochronology Centre (LGC) at University College London.

The data come from field areas on all seven continents. They were
acquired by LA-ICP-MS, on either an Agilent 7700x (old samples) or an
Agilent 7900 (recent samples) instrument, which was coupled to a New
Wave NWR193 excimer laser. Analytical conditions varied slightly
between different samples but generally used a 25 or 35 micron laser
spot, Ple\v{s}ovice zircon as a primary standard \citep{slama2008}
and, in most cases, GJ-1 zircon as a secondary standard
\citep{jackson2004}. Data reduction was done with GLITTER
\citep{griffin2008}. The data were not subjected to any common Pb
correction or other filters, apart from a visual inspection to remove
the most extreme outliers ($<1\%$ of the data).

Figure~\ref{fig:KDE} shows the frequency distribution of the complete,
unfiltered dataset as a kernel density estimate. The
\textsuperscript{207}Pb/\textsuperscript{235}U,
\textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
spectra all look similar. However, the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age distribution
deviates from the other three chronometers. It reduces the prominence
of the young age components, and inflates the old end of the age
spectrum.

@@

Figure~\ref{fig:KDEs} applies the six discordance filters to this
database. In order to emphasise the difference between the six
discordance definitions whilst treating them on an equal footing, each
of the filters was adjusted until half of the data were removed. This
was achieved by discordance cutoffs of $d_t$ = 51~Myr, $d_r = 0.0618$,
$d_{p} = 0.084$, $d_{sk} = 0.00225$, $d_a = 0.0219$ and $d_c =
0.0297$.

There are noticeable differences between the density estimates.  As
expected from the theoretical considerations laid out in
Sections~\ref{sec:discordance1} and \ref{sec:discordance2}, the
relative age filter greatly suppresses the younger age components
($<1.5$~Ga) relative to the older parts of the age spectrum
($>1.5$~Ga). The \citet{stacey1975} filter has the opposite effect.
It suppresses the Archaean age component by $\sim$20\% whilst further
increasing the prominence of the Neoproterozoic and Phanerozoic modes.

The discordance definitions based on the absolute age difference and
logratio distances have a comparatively minor effect on the shape of
the age spectrum. All six discordance filters reduce the difference
between the \textsuperscript{207}Pb/\textsuperscript{206}Pb and
concordia age spectra except for the p-value filter, which actually
exacerbates the problem.

If the degree of similarity between the four chronometers is taken as
a measure of their success, then the concordia distance filter ($d_c$)
is the most successful discordance criterion. Apart from a minor
difference at the youngest end of the age spectrum, the four different
chronometers yield virtually identical age distributions after passing
through this filter.

@@

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the six filters. In real applications, less
stringents discordance filters are usually applied. As mentioned in
the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=83.5$~Ma, a Stacey-Kramers filter with
$d_{sk}=0.381$\%, a perpendicular Aitchison filter with $d_a=3.61$\%,
and a concordia distance filter with $d_c=3.67$\%.

@@

\begin{table}
  \begin{tabular}{c|cccc}
    $d_r$ & $d_t$ & $d_{sk}$ & $d_a$ & $d_c$ \\ \hline
    -10 & -60.4 & -0.287 & -2.59 & -2.62 \\
    -5 & -38.8 & -0.175 & -1.67 & -1.69 \\
    -4 & -32.3 & -0.143 & -1.39 & -1.41 \\
    -3 & -25 & -0.108 & -1.08 & -1.09 \\
    -2 & -17.3 & -0.0681 & -0.747 & -0.758 \\
    -1 & -8.33 & -0.0328 & -0.356 & -0.361 \\
    0 & 0 & 0 & 0 & 0 \\
    1 & 9.31 & 0.0367 & 0.395 & 0.406 \\
    2 & 18.5 & 0.0773 & 0.795 & 0.816 \\
    3 & 28.1 & 0.117 & 1.2 & 1.23 \\
    4 & 36.7 & 0.151 & 1.58 & 1.6 \\
    5 & 43.9 & 0.189 & 1.9 & 1.94 \\
    10 & 83.5 & 0.381 & 3.61 & 3.67 \\
    15 & 117 & 0.538 & 5.02 & 5.16 \\
    20 & 151 & 0.736 & 6.54 & 6.7 \\
    25 & 187 & 0.949 & 8.01 & 8.24 \\
    30 & 230 & 1.18 & 9.93 & 10.2 \\
    40 & 328 & 1.74 & 14.2 & 14.8 \\
    50 & 463 & 2.47 & 20.3 & 21.6
  \end{tabular}
  \caption{Conversion table for the different discordance filters,
    constructed using the test data. All discordance values are
    expressed as \%, except for $d_t$, which is expressed in Ma. This
    table allows the reader to select a discordance cutoff that
    removes the same fraction of their data as the relative age cutoff
    ($d_r$) that they may have applied in the past. }
  \label{tab:conversiontable}
\end{table}

@@

\noindent Loading the test data into memory:

\begin{verbatim}
UPb <- read.data('data.csv',method='U-Pb',format=2)
\end{verbatim}

@@

\noindent If the dataset includes \textsuperscript{204}Pb (which is
not the case for the test data), then we can also apply a discordance
filter \textit{after} the common Pb correction. For example:

@@

\authorcontribution{This is a single author publication.} %% this section is mandatory

>>>>

@@

\title{On the treatment of discordant data in detrital zircon U--Pb
  geochronology}

@@

\begin{abstract}
  Detrital zircon U--Pb geochronology is a staple of sedimentary
  provenance analysis and crustal evolution studies. Calculating
  detrital age spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This situation is expected to
  worsen in the future as a result of increasing analytical precision,
  which reveals ever smaller deviations from isotopic concordance.
  The analysis of discordant U--Pb data involves two largely arbitrary
  decisions.

@@

  Applying these six discordance filters to a 10,000-grain dataset of
  detrital zircon U--Pb compositions reveals that: (i) the relative
  age discordance filter tends to suppress the young ($<1.5$~Ga) age
  components in U--Pb age spectra, whilst inflating the older
  ($>1.5$~Ga) age components; (ii) the absolute age difference and
  Stacey-Kramers discordance filters are more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it yields age spectra that are
  identical for the \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb clocks.  All the
  methods presented in this paper have been implemented in the
  \texttt{IsoplotR} geochronology toolbox.
\end{abstract}

@@

This paper addresses both of these issues. Section~\ref{sec:whichage}
shows that single-grain concordia ages can be used to avoid the
arbitrary cutoff between the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb
methods. Section~\ref{sec:discordance1} compares and contrasts
existing discordance filters based on age disparity and p-values. It
shows that the p-value definition hurts both the accuracy and
precision of detrital geochronology. Section~\ref{sec:discordance2}
proposes three new ways to quantify discordance, based directly on
U--Pb compositions rather than the ages calculated therefrom. Two of
these new definitions are based on a logratio distance that is widely
used in compositional data analysis. Finally,
Section~\ref{sec:application} applies the six discordance filters to a
compilation of detrital zircon U--Pb data. Although the true age
distribution of this dataset is unknowable, the results suggest that
the logratio based discordance filters produce the most accurate
results.

@@

The U--Pb method is based on three separate chronometers:
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb. The half-life of
\textsuperscript{235}U is more than six times shorter than that of
\textsuperscript{238}U, and \textsuperscript{235}U is more than 100
times less abundant than \textsuperscript{238}U. For these two
reasons, very little \textsuperscript{207}Pb has been produced during
the last billion years of Earth history compared to
\textsuperscript{206}Pb. Consequently, the
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb methods are less
precise than the \textsuperscript{206}Pb/\textsuperscript{238}U method
during the Phanerozoic and Neoproterozoic.\\

However, during earlier stages of Earth's history,
\textsuperscript{235}U was significantly more abundant than it is
today. The \textsuperscript{238}U/\textsuperscript{235}U ratio was
$\sim$60 at 1Ga, $\sim$26 at 2Ga, $\sim$11 at 3Ga, and $\sim$5 at 4Ga.
Due to the greater abundance of \textsuperscript{235}U in this past,
and because it decays much faster than \textsuperscript{238}U, the
precision of the \textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb clocks exceeds that of
the \textsuperscript{206}Pb/\textsuperscript{238}U method during the
Palaeoproterozoic and Archaean. The gradual shift in sensitivity
between the two chronometers is visible in the slope of a
Tera-Wasserburg concordia line, which is steep at old ages (high
\textsuperscript{207}Pb/\textsuperscript{206}Pb gradient w.r.t. time)
and shallow at young ages (low
\textsuperscript{238}U/\textsuperscript{206}Pb gradient
w.r.t. time).\\

Most published detrital zircon U--Pb studies switch from
\textsuperscript{206}Pb/\textsuperscript{238}U to
\textsuperscript{207}Pb/\textsuperscript{206}Pb at some point during
the Proterozoic. Unfortunately there are two problems with such a
switch. First, it requires the selection of a discrete discordance
cutoff between the two methods. If this cutoff differs between two
studies (which it often does), then this complicates the
intercomparison of their respective age spectra. Second, the sudden
switch between the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb clocks is often marked
by a discrete step in the age spectrum \citep{puetz2018}. This step is
entirely artificial and obscures any geologically significant events
that might occur around the same time.\\

Both of these problems can be solved by using `hybrid' concordia ages
instead of `pure' \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages. Concordia ages
are defined by \citet{ludwig1998} as the `most likely' (in a
statistical sense) U--Pb age given the isotopic ratio composition and
its analytical uncertainty (Figure~\ref{fig:concordia}).  Let $r_{75}$
and $r_{68}$ be the measured
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{206}Pb/\textsuperscript{238}U ratios, respectively,
and let $\sigma[r_{68}]^2$, $\sigma[r_{76}]^2$,
$\sigma[r_{68},r_{76}]$ be their (co)variances.  Then the concordia
age $t_c$ is obtained by numerically minimising the sum of squares
$S$:
\begin{equation}
  S = \left[
    \begin{array}{c}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1
    \end{array}
    \right]^T\left[
    \begin{array}{cc}
      \sigma[r_{75}]^2 & \sigma[r_{75},r_{68}] \\
      \sigma[r_{75},r_{68}] & \sigma[r_{68}]^2
    \end{array}
    \right]^{-1}
  \left[
    \begin{array}{c}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1      
    \end{array}
    \right]
  \label{eq:S}
\end{equation}

@@

\subsection{Discordance filters: old definitions}\label{sec:discordance1}

@@

Detrital zircon U--Pb data can be filtered by removing all
measurements whose discordance values exceed a certain threshold
value. Typical cutoff values for $d_r$ are 10--30\%
\citep{gehrels2011}, whereas $d_p$ is generally set to 5\%
\citep{spencer2016}. Different discordance criteria produce different
U--Pb age spectra. For example, a relative age cutoff will
preferentially remove young grains whereas an absolute age cutoff is
more likely to remove old grains (Figure~\ref{fig:agediscordance}).\\

@@

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is 1\% discordant. If this grain were analysed by
LA-ICP-MS with an analytical precision of 2\%, say, then it would pass
the chi-square test and be accepted as being concordant. However, if
that same grain were analysed by TIMS with a precision of 0.2\%, then
the p-value criterion would reject it as being discordant. It seems
fundamentally unfair that an imprecise analytical method would be
favoured over a precise one (Figure~\ref{fig:TIMSvsLAICPMS}). This is
a pertinent problem because technical innovations are increasing the
precision of all analytical approaches to U--Pb geochronology.  As
precision improves, so does the ability to detect even small degrees
of discordance. Using the p-value criterion, there may come a time
when no detrital zircon passes this filter.\\

@@

\subsubsection{Discordance filters: new definitions}\label{sec:discordance2}

@@

\noindent where $[{}^{238}U/{}^{206}Pb]^\ast$ is the
\textsuperscript{238}U/\textsuperscript{206}Pb-ratio of the
intersection between concordia and a straight line connecting the
\textsuperscript{238}U/\textsuperscript{206}Pb--\textsuperscript{207}Pb/\textsuperscript{206}Pb
measurement and the inferred mantle composition
(Figure~\ref{fig:SK}).\\

The common Pb definition of discordance is more forgiving for young
grains than it is for old ones. Importantly, if the discordance is
caused by common Pb, then the
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{206}Pb and concordia age
estimates are all negatively biased with respect to the true
age. However this bias can be removed by applying a common-Pb
correction \emph{after} the data have been filtered.\\

Each discordia definition that we have studied thus far is expressed
in different units. For the absolute age definition, degrees of
discordance are expressed in units of time (ranging from 0 to
4.5~Ga). The relative age definition uses fractions of time (ranging
from $-\infty$ to 1). The p-value definition expresses discordance in
terms of probability (ranging from 0 to 1). And the \citet{stacey1975}
definition uses fractions of ratios (ranging from $-\infty$ to
1). None of these scales is particularly intuitive or natural. They
certainly do not match the usual definition of \emph{distance} in the
geographical sense of the word.\\

To address this issue, it is useful to subject the U--Pb isotopic
ratio data to a logarithmic transformation. So instead of analysing
our data on a conventional Tera-Wasserburg concordia diagram, all
calculations can be done in
log(\textsuperscript{207}Pb/\textsuperscript{206}Pb) vs.
log(\textsuperscript{238}U/\textsuperscript{206}Pb) space. The
advantage of this transformation is that it produces values that are
free to range from $-\infty$ to $+\infty$. Within this infinite
dataspace, the Euclidean distance metric can be safely applied.\\

\begin{figure}
  \includegraphics[width=\textwidth]{Aitchison.pdf}
  \caption{Illustration of the two logratio distance definitions of
    discordance. $d_a$ is the perpendicular Aitchison distance from
    the measured logratio to the concordia line. $d_c$ is the
    Aitchison distance measured along a line connecting the measured
    value and the concordia composition.
  }
  \label{fig:aitchison}
\end{figure}

@@

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Era, when the U--Pb method is most reliable
(Figure~\ref{fig:dperp}).\\

The perpendicular Aitchison distance criterion does not take into
account the analytical precision of the isotopic measurements.  To
address this issue, we can also measure the Aitchison distance along a
line connecting the measured logratio and the maximum likelihood
composition on the concordia line:
\begin{equation}
  d_c = \sgn[t_{76}-t_{68}] \sqrt{ dx(t_c)^2 + dy(t_c)^2 }
  \label{eq:da}
\end{equation}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{TW-option-4.pdf}
  \caption{The $d_a$ criterion produces a parallel band to the
    (logarithmic) Tera-Wasserburg concordia line. The dashed line is
    plotted at an Aitchison distance of 0.1 from concordia, marking
    a 10\% discordance cutoff. The bottom panel presents the age
    discordance along this line. This shows that the $d_a$ criterion
    is lenient for Phanerozoic and Archaean age estimates, and more
    strict for Proterozoic grains.
  }
  \label{fig:dperp}
\end{figure}

@@

It is difficult to ascertain the mechanism causing discordance in any
particular zircon grain. Therefore, it is unclear which of the
definitions in Sections~\ref{sec:discordance1} and
\ref{sec:discordance2} is `correct'. All we can is do is apply the
methods to real samples and investigate their outcomes. This Section
will apply the six discordance filters to a compilation of 10,000
detrital zircon U--Pb analyses that were acquired at the London
Geochronology Centre (LGC) at University College London.\\

The data come from field areas on all seven continents. They were
acquired by LA-ICP-MS, on either an Agilent 7700x (old samples) or an
Agilent 7900 (recent samples) instrument, which was coupled to a New
Wave NWR193 excimer laser. Analytical conditions varied slightly
between different samples but generally used a 25 or 35 micron laser
spot, Ple\v{s}ovice zircon as a primary standard \citep{slama2008}
and, in most cases, GJ-1 zircon as a secondary standard
\citep{jackson2004}. Data reduction was done with GLITTER
\citep{griffin2008}. The data were not subjected to any common Pb
correction or other filters, apart from a visual inspection to remove
the most extreme outliers ($<1\%$ of the data).\\

Figure~\ref{fig:KDE} shows the frequency distribution of the complete,
unfiltered dataset as four kernel density estimates
\citep{vermeesch2012b}. The
\textsuperscript{207}Pb/\textsuperscript{235}U,
\textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
spectra all look similar. However, the
\textsuperscript{207}Pb/\textsuperscript{206}Pb age distribution
deviates from the other three chronometers. It reduces the prominence
of the young age components, and inflates the old end of the age
spectrum.\\

@@

Figure~\ref{fig:KDEs} applies the six discordance filters to this
database. In order to emphasise the difference between the six
discordance definitions whilst treating them on an equal footing, each
of the filters was adjusted until half of the data were removed. This
was achieved by discordance cutoffs of $d_t$ = 51~Myr, $d_r = 0.0618$,
$d_{p} = 0.088$, $d_{sk} = 0.00218$, $d_a = 0.0219$ and $d_c =
0.0324$.\\

There are significant differences between the density estimates.  As
expected from the theoretical considerations laid out in
Sections~\ref{sec:discordance1} and \ref{sec:discordance2}, the
relative age filter greatly suppresses the younger age components
($<1.5$~Ga) relative to the older parts of the age spectrum
($>1.5$~Ga). The \citet{stacey1975} filter has the opposite effect.
It suppresses the Archaean age component by $\sim$20\% whilst further
increasing the prominence of the Neoproterozoic and Phanerozoic
modes.\\

The discordance definitions based on the absolute age difference and
logratio distances have a comparatively minor effect on the shape of
the age spectrum. All six discordance filters reduce the difference
between the \textsuperscript{207}Pb/\textsuperscript{206}Pb and
concordia age spectra except for the p-value filter, which actually
exacerbates the problem.\\

If the degree of similarity between the four chronometers is taken as
a measure of their success, then the concordia distance filter ($d_c$)
is the most successful discordance criterion. Apart from a minor
difference at the youngest end of the age spectrum, the four different
chronometers yield virtually identical age distributions after passing
through this filter.\\

@@

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the six filters. In real applications, less
stringents discordance filters are usually applied. As mentioned in
the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=83.5$~Ma, a Stacey-Kramers filter with
$d_{sk}=0.371$\%, a perpendicular Aitchison filter with $d_a=3.61$\%,
and a concordia distance filter with $d_c=$.\\

@@

\begin{table}
  \begin{minipage}[c]{0.45\textwidth}
  \begin{tabular}{c|cccc}
    $d_r$ & $d_t$ & $d_{sk}$ & $d_a$ & $d_c$ \\ \hline
-10 & -60.4 & -0.287 & -2.59 & -2.62 \\
-5 & -38.8 & -0.175 & -1.67 & -1.69 \\
-4 & -32.3 & -0.143 & -1.39 & -1.41 \\
-3 & -25 & -0.108 & -1.08 & -1.09 \\
-2 & -17.3 & -0.0681 & -0.747 & -0.758 \\
-1 & -8.33 & -0.0328 & -0.356 & -0.361 \\
0 & 0 & 0 & 0 & 0 \\
1 & 9.31 & 0.0367 & 0.395 & 0.406 \\
2 & 18.5 & 0.0773 & 0.795 & 0.816 \\
3 & 28.1 & 0.117 & 1.2 & 1.23 \\
4 & 36.7 & 0.151 & 1.58 & 1.6 \\
5 & 43.9 & 0.189 & 1.9 & 1.94 \\
10 & 83.5 & 0.381 & 3.61 & 3.67 \\
15 & 117 & 0.538 & 5.02 & 5.16 \\
20 & 151 & 0.736 & 6.54 & 6.7 \\
25 & 187 & 0.949 & 8.01 & 8.24 \\
30 & 230 & 1.18 & 9.93 & 10.2 \\
40 & 328 & 1.74 & 14.2 & 14.8 \\
50 & 463 & 2.47 & 20.3 & 21.6 \\
  \end{tabular}
  \end{minipage}
  \begin{minipage}[c]{0.5\textwidth}
  \caption{Conversion table for the different discordance filters,
    constructed using the test data. All discordance values are
    expressed as \%, except for $d_t$, which is expressed in Ma. This
    table allows the reader to select a discordance cutoff that
    removes the same fraction of their data as the relative age cutoff
    ($d_r$) that they may have applied in the past. }
  \label{tab:conversiontable}
  \end{minipage}
\end{table}

@@

\noindent Loading the test data (see Supplementary Information) into
memory:

\begin{verbatim}
UPb <- read.data('LGCdata.csv',method='U-Pb',format=2)
\end{verbatim}

@@

\noindent If the dataset includes \textsuperscript{204}Pb (which is not
the case for the test data), then we can also apply the discordance
filter \textit{after} applying the common Pb correction. For example:

@@

\authorcontribution{} %% this section is mandatory
